Wall-to-wall mapping of peat depth from Lidar terrain and airborne radiometrics in Norwegian landscapes

### 2024-10-29 ms/ms.Rmd
Model interpretation with noncollinear features Added pdp Harmonized modeling across sites Ms: model approach, performance, interpretation

We created the same suite of 25 quantitative peat depth predictors for both sites (Table \@ref(tab:preds)).
All quantitative predictors were derived either from an airborne radiometric survey or from a DTM.
From the radiometric surveys we simply used the four variables produced by the surveyors (Geological Survey of Norway): ground concentration of Potassium, Thorium, Uranium, as well as total count.
From the DTMs we calculated several land surface parameters, ranging from simple terrain indices to more complex geomorphometric and hydrological variables [@maxwellLandsurfaceParametersSpatial2022].
```{r preds, echo=FALSE, message=FALSE}
df <- readr::read_csv("tables/predictors.csv")
knitr::kable(df, booktabs = FALSE, caption = "Quantitative predictors of peat depth.")
```

@@

@@

We used the *terra* R package to calculate from the DTMs: slope, topographic position index (difference from mean of eight neighbors), terrain ruggedness index (mean of absolute differences from eight neighbors), and roughness (range in the nine-cell neighborhood).
These variables were derived at two scales to produce different predictors; we either calculated from \unit{1\,m} DTM resolution and then aggregated, or aggregated to \unit{10\,m} DTM resolution and then calculated the indices.
This kind of multiscale feature engineering of land surface parameters has been found to improve machine learning predictions of soil properties [@millerImpactMultiscalePredictor2015; @dornikOptimalScalingPredictors2022; @newmanAssessingSpatiallyHeterogeneous2023].
We chose \unit{1\,m} and \unit{10\,m} resolutions because we know that peat depth tends to vary at fine scales in Norway [@maxwellLandsurfaceParametersSpatial2022].
We also calculated the the Multi-Resolution Valley Bottom Flatness index, which indicates the degree of valley bottom flatness at a given location via a multiscale algorithm [@gallantMultiresolutionIndexValley2003].
We calculated this index in SAGA GIS [v.9.3.2, @conradSystemAutomatedGeoscientific2015] with default parameters.

@@

In addition to the quantitative predictors described above, we prepared one categorical predictor --- peat depth class --- from a historical national map dataset called *DMK* [@ahlstromAR5Klassifikasjonssystem2019].
The DMK peat depth classes are: < \unit{1\,m} (*shallow*), > \unit{1\,m} (*deep*), and *unknown*.
This dataset stems from field measurements made in 1964--2007 as part of a wider land cover mapping in Norway.
The mapping was primarily for identifying agricultural and silvicultural resources, so it covers productive areas below the tree line and has greater coverage of peat depth class in peatlands judged to be arable or afforestable [@bjordalMarkslagsklassifikasjonOkonomiskKartverk2007; @ahlstromAR5Klassifikasjonssystem2019].
Mappers generally assigned peat depth classes to polygons of at least \unit{0.5\,ha}, although delineating polygons down to \unit{0.2\,ha} was allowed if peat depth showed a "particularly marked difference" [@bjordalMarkslagsklassifikasjonOkonomiskKartverk2007].
We rasterized the peat depth class attribute to our \unit{10\,m} grid.

@@

@@

For each site we followed the same modelling workflow.
First we trained a RF to predict peat depth based on our 25 quantitative predictors, to simulate applications where radiometric and terrain data are available, but no maps of peat depth.
Next we trained a RF with all 26 predictors (DMK peat depth class included), to fully leverage the relevant available data and check the added value of the national DMK dataset.
Finally, we fit a simple linear model to only DMK depth class, to provide a fair comparison between the accuracy of the RF models and the existing national map of peat depth calibrated on the same data.
We implemented models in the *tidymodels* framework in R [@kuhnTidymodelsCollectionPackages2020], with the *ranger* R package for RFs [v.0.16, @wrightRangerFastImplementation2017].
RFs were fit with 1000 trees, minimum node size of 5, and the number of predictors randomly sampled at each split was the square root of the total number of predictors (*ranger* default).
We did not tune these hyperparameters because RFs are relatively insensitive to tuning [@probstHyperparametersTuningStrategies2019], and because it would require nested spatial cross-validation to prevent data leakage [@schratzHyperparameterTuningPerformance2019].

The performance of digital soil mapping must be evaluated with reference to a specific purpose [i.e., map vs. model validation, interpolation vs. extrapolation, @robertsCrossvalidationStrategiesData2017; @milaNearestNeighbourDistance2022], and here we aimed to evaluate maps covering the study areas.
In the absence of additional field work to collect a design-based independent validation set, we used a spatial cross-validation scheme to evaluate model performance [@wadouxSpatialCrossvalidationNot2021; @meyerMachineLearningbasedGlobal2022].
Specifically, we used k-Means Nearest Neighbor Distance Matching (kNNDM), which creates cross-validation folds that mimic the spatial prediction task that is defined as the goal [@linnenbrinkKNNDMCVKfold2024].
In particular, kNNDM looks for the spatial assignment of training data to folds that minimizes the difference between two distributions: nearest neighbor distances between training and test locations in the cross-validation, and nearest neighbor distances between training and prediction locations for the model.
That way, the spatial separation between folds is similar to the separation between training and prediction locations --- which increases the quality of the map accuracy estimate [@linnenbrinkKNNDMCVKfold2024].
For spatially clustered training data, this approach strikes a balance between the risk of optimistic metrics from random cross-validation and the risk of pessimistic metrics from other forms of spatial cross-validation [@wadouxSpatialCrossvalidationNot2021].
We implemented the kNNDM with the *CAST* R package [v.1.0.2, @meyerCASTPackageTraining2024], setting prediction locations to all AR5 mire cells in the study area, and choosing a number of folds (5--20) that produced the best match between the two NND distributions.
From the cross-validation we quantified *root mean squared error* (accuracy), *R^2^* (correlation), and *Lin's concordance correlation coefficient* (accuracy + correlation).

We quantified global variable importance for the RFs with the 25 quantitative predictors.
Global variable importance measures the influence of a given predictor on the output of the model, aggregated across all locations.
We calculated variable importance with the *vip* R package (v.0.4.1), by three different methods: *FIRM*, *permutation*, and *Shapley* [@greenwellVariableImportancePlots2020].
*Permutation* values were aggregated from ten iterations with RMSE as the performance measure.

>>>>

We created the same suite of peat depth predictors for both sites (25 continuous and 1 categorical; Table \@ref(tab:preds)).
All continuous predictors were derived either from an airborne radiometric survey or from a DTM.
From the radiometric surveys we simply used the four variables produced by the surveyors (Geological Survey of Norway): ground concentration of Potassium, Thorium, Uranium, as well as total count.
From the DTMs we calculated several land surface parameters, ranging from simple terrain indices to more complex geomorphometric and hydrological variables [@maxwellLandsurfaceParametersSpatial2022].
The categorical predictor was peat depth class, from a national map dataset.
Complete descriptions of predictors follow below.
```{r preds, echo=FALSE, message=FALSE}
df <- readr::read_csv("tables/predictors.csv")
knitr::kable(df, booktabs = FALSE, caption = "Candidate predictors of peat depth.")
```

### Radiometric

@@

### Terrain

@@

We used the *terra* R package to calculate from the DTMs: slope, topographic position index (difference from mean of eight neighbors), terrain ruggedness index (mean of absolute differences from eight neighbors), and roughness (range in the nine-cell neighborhood).
These variables were derived at two scales to produce different predictors; we either calculated from \unit{1\,m} DTM resolution and then aggregated, or aggregated to \unit{10\,m} DTM resolution and then calculated the indices.
This kind of multiscale feature engineering of land surface parameters has been found to improve machine learning predictions of soil properties [@millerImpactMultiscalePredictor2015; @dornikOptimalScalingPredictors2022; @newmanAssessingSpatiallyHeterogeneous2023].
We know that peat depth tends to vary at fine scales in Norway, which is why we chose \unit{1\,m} and \unit{10\,m} resolutions [@maxwellLandsurfaceParametersSpatial2022].
We also calculated the the Multi-Resolution Valley Bottom Flatness index, which indicates the degree of valley bottom flatness at a given location via a multiscale algorithm [@gallantMultiresolutionIndexValley2003].
We calculated this index in SAGA GIS [v.9.3.2, @conradSystemAutomatedGeoscientific2015] with default parameters.

@@

### Peat depth class

We prepared one categorical predictor --- peat depth class --- from a historical national map dataset called *DMK* [@ahlstromAR5Klassifikasjonssystem2019].
The DMK peat depth classes are: < \unit{1\,m} (*shallow*), > \unit{1\,m} (*deep*), and *unknown*.
This dataset stems from field measurements made in 1964--2007 as part of a wider land cover mapping in Norway.
The mapping was primarily for identifying agricultural and silvicultural resources, so it covers productive areas below the tree line and has greater coverage of peat depth class in peatlands judged to be arable or afforestable [@bjordalMarkslagsklassifikasjonOkonomiskKartverk2007; @ahlstromAR5Klassifikasjonssystem2019].
Mappers generally assigned peat depth classes to polygons of at least \unit{0.5\,ha}, although delineating polygons down to \unit{0.2\,ha} was allowed if peat depth showed a "particularly marked difference" [@bjordalMarkslagsklassifikasjonOkonomiskKartverk2007].
We rasterized the peat depth class attribute to our \unit{10\,m} grid.

@@

### Modelling approach

@@

We implemented models in the *tidymodels* framework in R [@kuhnTidymodelsCollectionPackages2020], with the *ranger* R package for RFs [v.0.16, @wrightRangerFastImplementation2017].
RFs were fit with 1000 trees, minimum node size of 5, and the number of predictors randomly sampled at each split was the square root of the total number of predictors (*ranger* default).
We did not tune these hyperparameters because RFs are relatively insensitive to tuning [@probstHyperparametersTuningStrategies2019], and because it would require nested spatial cross-validation to prevent data leakage [@schratzHyperparameterTuningPerformance2019].

### Model performance

For both sites we compared the performance of models with five different configurations of three groups of predictors. 
Specifically, we trained models with:

  1. only DMK peat depth class (2 binary indicator variables from 3 levels)
  2. only terrain (21 continuous variables)
  3. terrain and DMK peat depth class (23 variables)
  4. radiometric and terrain predictors (25 continuous variables)
  5. all predictors (26 variables)

These different configurations simulate different scenarios of data availability that are common in Norway or internationally.
Comparing the different configurations allowed us to isolate the added value of each of the predictor groups.
We did not explore configurations comprising radiometrics without terrain, because Lidar terrain surveys typically precede airborne radiometric surveys.
The models with only DMK peat depth class were simple linear models rather than RFs, and served to provide a fair comparison between the accuracy of the RF models and the existing national map of peat depth, calibrated on the same data.

The performance of digital soil mapping must be evaluated with reference to a specific purpose [i.e., map vs. model validation, interpolation vs. extrapolation, @robertsCrossvalidationStrategiesData2017; @milaNearestNeighbourDistance2022], and here we aimed to evaluate maps of peat depth across the study areas.
In the absence of additional field work to collect a design-based independent validation set, we used a spatial cross-validation scheme to evaluate model performance [@wadouxSpatialCrossvalidationNot2021; @meyerMachineLearningbasedGlobal2022].
Specifically, we used k-Means Nearest Neighbor Distance Matching (kNNDM), which creates cross-validation folds that mimic the spatial prediction task that is defined as the goal [@linnenbrinkKNNDMCVKfold2024].
In particular, kNNDM looks for the spatial assignment of training data to folds that minimizes the difference between two distributions: nearest neighbor distances between training and test locations in the cross-validation, and nearest neighbor distances between training and prediction locations for the model.
That way, the spatial separation between folds is similar to the separation between training and prediction locations --- which increases the quality of the map accuracy estimate [@linnenbrinkKNNDMCVKfold2024].
For spatially clustered training data, this approach strikes a balance between the risk of optimistic metrics from random cross-validation and the risk of pessimistic metrics from other forms of spatial cross-validation [@wadouxSpatialCrossvalidationNot2021].
We implemented the kNNDM with the *CAST* R package [v.1.0.2, @meyerCASTPackageTraining2024], setting prediction locations to all AR5 mire cells in the study area, and choosing a number of folds (5--20) that produced the best match between the two NND distributions.
From the cross-validation we quantified *Root Mean Squared Error* (accuracy, original scale), *R^2^* (correlation, standardized scale), and *Lin's concordance correlation coefficient* (accuracy and correlation, standardized scale).

Uncertainty.

Extrapolation.

### Model interpretation

We quantified global variable importance and examined partial dependence plots for the best-performing predictor configuration at each site.
Global variable importance measures the influence of a given predictor on the output of the model, aggregated across all locations.
Partial dependence plots depict the shape of the fitted relationship between a given predictor and the outcome.
Both are useful for understanding the mechanisms behind the model's predictions and the roles of the predictors in the model.

For both sites, we interpreted a model trained on a non-collinear subset of variables from the best performing predictor configuration --- because correlation between predictors degrades variable importance measures [@stroblConditionalVariableImportance2008; @biauRandomForestGuided2016] and can produce misleading visualizations of predictor-outcome relationships [@biecekExplanatoryModelAnalysis2021; @dwivediExplainableAIXAI2023].
Specifically, we eliminated variables from the best performing predictor configuration to obtain a set with no pairwise Pearson correlation above 0.7.
Thus, highly-correlated sets of predictors are represented by a single variable for the purposes of model interpretation.

We calculated variable importance with the *vip* R package (v.0.4.1), by three different methods: *FIRM*, *permutation*, and *Shapley* [@greenwellVariableImportancePlots2020].
*FIRM* values measure the flatness of the partial dependence plot, *permutation* values measure the decrease in model performance when the predictor is permuted, and *Shapley* values are aggregated from local, game-theoretical measures of variable importance [@greenwellVariableImportancePlots2020]. 
*Permutation* values were obtained from ten iterations, with RMSE as the performance measure.

We calculated partial dependence with the *pdp* R package [v.0.8.1, @greenwellPdpPackageConstructing2017].
For the top six most important variables, we plotted both partial dependence and individual conditional expectation (ICE), to show the average effect of the predictor on the outcome and the variation in the effect across observations, repectively [@goldsteinPeekingBlackBox2015].
Non-parallel ICE lines indicate the presence of interactions between predictors.

### 2024-10-29 ms/ms.tex
Model interpretation with noncollinear features Added pdp Harmonized modeling across sites Ms: model approach, performance, interpretation

We created the same suite of 25 quantitative peat depth predictors for both sites (Table \ref{tab:preds}).
All quantitative predictors were derived either from an airborne radiometric survey or from a DTM.
From the radiometric surveys we simply used the four variables produced by the surveyors (Geological Survey of Norway): ground concentration of Potassium, Thorium, Uranium, as well as total count.
From the DTMs we calculated several land surface parameters, ranging from simple terrain indices to more complex geomorphometric and hydrological variables \citep{maxwellLandsurfaceParametersSpatial2022}.

@@

\caption{\label{tab:preds}Quantitative predictors of peat depth.}
\centering
\begin{tabular}[t]{l|l}
\hline
name & description\\
\hline
radK & Potassium ground concentration\\
\hline
radTh & Thorium ground concentration\\
\hline
radU & Uranium ground concentration\\
\hline
radTC & Total count of gamma radiation\\
\hline
elevation & Mean elevation\\
\hline
slope1m & Mean of 1 m slope\\
\hline
TPI1m & Mean of 1 m topographic position index\\
\hline
TRI1m & Mean of 1 m terrain ruggedness index\\
\hline
roughness1m & Mean of 1 m roughness\\
\hline
slope10m & 10 m slope\\
\hline
TPI10m & 10 m topographic position index\\
\hline
TRI10m & 10 m terrain ruggedness index\\
\hline
roughness10m & 10 m roughness\\
\hline
MRVBF & Multi-resolution valley bottom flatness\\
\hline
TWI5m & Mean of 5 m topographic wetness index\\
\hline
TWI10m & 10 m topographic wetness index\\
\hline
TWI20m & Bilinear interpolation of 20 m topographic wetness index\\
\hline
TWI50m & Bilinear interpolation of 50 m topographic wetness index\\
\hline
DTW2500 & Depth-to-water index, flow initiation area of 0.25 ha\\
\hline
DTW5000 & Depth-to-water index, flow initiation area of 0.5 ha\\
\hline
DTW10000 & Depth-to-water index, flow initiation area of 1 ha\\
\hline
DTW20000 & Depth-to-water index, flow initiation area of 2 ha\\
\hline
DTW40000 & Depth-to-water index, flow initiation area of 4 ha\\
\hline
DTW80000 & Depth-to-water index, flow initiation area of 8 ha\\
\hline
DTW160000 & Depth-to-water index, flow initiation area of 16 ha\\
\hline
\end{tabular}
\end{table}

@@

@@

We used the \emph{terra} R package to calculate from the DTMs: slope, topographic position index (difference from mean of eight neighbors), terrain ruggedness index (mean of absolute differences from eight neighbors), and roughness (range in the nine-cell neighborhood).
These variables were derived at two scales to produce different predictors; we either calculated from \unit{1\,m} DTM resolution and then aggregated, or aggregated to \unit{10\,m} DTM resolution and then calculated the indices.
This kind of multiscale feature engineering of land surface parameters has been found to improve machine learning predictions of soil properties \citep{millerImpactMultiscalePredictor2015, dornikOptimalScalingPredictors2022, newmanAssessingSpatiallyHeterogeneous2023}.
We chose \unit{1\,m} and \unit{10\,m} resolutions because we know that peat depth tends to vary at fine scales in Norway \citep{maxwellLandsurfaceParametersSpatial2022}.
We also calculated the the Multi-Resolution Valley Bottom Flatness index, which indicates the degree of valley bottom flatness at a given location via a multiscale algorithm \citep{gallantMultiresolutionIndexValley2003}.
We calculated this index in SAGA GIS \citep[v.9.3.2,][]{conradSystemAutomatedGeoscientific2015} with default parameters.

@@

In addition to the quantitative predictors described above, we prepared one categorical predictor --- peat depth class --- from a historical national map dataset called \emph{DMK} \citep{ahlstromAR5Klassifikasjonssystem2019}.
The DMK peat depth classes are: \textless{} \unit{1\,m} (\emph{shallow}), \textgreater{} \unit{1\,m} (\emph{deep}), and \emph{unknown}.
This dataset stems from field measurements made in 1964--2007 as part of a wider land cover mapping in Norway.
The mapping was primarily for identifying agricultural and silvicultural resources, so it covers productive areas below the tree line and has greater coverage of peat depth class in peatlands judged to be arable or afforestable \citep{bjordalMarkslagsklassifikasjonOkonomiskKartverk2007, ahlstromAR5Klassifikasjonssystem2019}.
Mappers generally assigned peat depth classes to polygons of at least \unit{0.5\,ha}, although delineating polygons down to \unit{0.2\,ha} was allowed if peat depth showed a ``particularly marked difference'' \citep{bjordalMarkslagsklassifikasjonOkonomiskKartverk2007}.
We rasterized the peat depth class attribute to our \unit{10\,m} grid.

@@

@@

For each site we followed the same modelling workflow.
First we trained a RF to predict peat depth based on our 25 quantitative predictors, to simulate applications where radiometric and terrain data are available, but no maps of peat depth.
Next we trained a RF with all 26 predictors (DMK peat depth class included), to fully leverage the relevant available data and check the added value of the national DMK dataset.
Finally, we fit a simple linear model to only DMK depth class, to provide a fair comparison between the accuracy of the RF models and the existing national map of peat depth calibrated on the same data.
We implemented models in the \emph{tidymodels} framework in R \citep{kuhnTidymodelsCollectionPackages2020}, with the \emph{ranger} R package for RFs \citep[v.0.16,][]{wrightRangerFastImplementation2017}.
RFs were fit with 1000 trees, minimum node size of 5, and the number of predictors randomly sampled at each split was the square root of the total number of predictors (\emph{ranger} default).
We did not tune these hyperparameters because RFs are relatively insensitive to tuning \citep{probstHyperparametersTuningStrategies2019}, and because it would require nested spatial cross-validation to prevent data leakage \citep{schratzHyperparameterTuningPerformance2019}.

The performance of digital soil mapping must be evaluated with reference to a specific purpose \citep[i.e., map vs.~model validation, interpolation vs.~extrapolation,][]{robertsCrossvalidationStrategiesData2017, milaNearestNeighbourDistance2022}, and here we aimed to evaluate maps covering the study areas.
In the absence of additional field work to collect a design-based independent validation set, we used a spatial cross-validation scheme to evaluate model performance \citep{wadouxSpatialCrossvalidationNot2021, meyerMachineLearningbasedGlobal2022}.
Specifically, we used k-Means Nearest Neighbor Distance Matching (kNNDM), which creates cross-validation folds that mimic the spatial prediction task that is defined as the goal \citep{linnenbrinkKNNDMCVKfold2024}.
In particular, kNNDM looks for the spatial assignment of training data to folds that minimizes the difference between two distributions: nearest neighbor distances between training and test locations in the cross-validation, and nearest neighbor distances between training and prediction locations for the model.
That way, the spatial separation between folds is similar to the separation between training and prediction locations --- which increases the quality of the map accuracy estimate \citep{linnenbrinkKNNDMCVKfold2024}.
For spatially clustered training data, this approach strikes a balance between the risk of optimistic metrics from random cross-validation and the risk of pessimistic metrics from other forms of spatial cross-validation \citep{wadouxSpatialCrossvalidationNot2021}.
We implemented the kNNDM with the \emph{CAST} R package \citep[v.1.0.2,][]{meyerCASTPackageTraining2024}, setting prediction locations to all AR5 mire cells in the study area, and choosing a number of folds (5--20) that produced the best match between the two NND distributions.
From the cross-validation we quantified \emph{root mean squared error} (accuracy), \emph{R\textsuperscript{2}} (correlation), and \emph{Lin's concordance correlation coefficient} (accuracy + correlation).

We quantified global variable importance for the RFs with the 25 quantitative predictors.
Global variable importance measures the influence of a given predictor on the output of the model, aggregated across all locations.
We calculated variable importance with the \emph{vip} R package (v.0.4.1), by three different methods: \emph{FIRM}, \emph{permutation}, and \emph{Shapley} \citep{greenwellVariableImportancePlots2020}.
\emph{Permutation} values were aggregated from ten iterations with RMSE as the performance measure.

>>>>

We created the same suite of peat depth predictors for both sites (25 continuous and 1 categorical; Table \ref{tab:preds}).
All continuous predictors were derived either from an airborne radiometric survey or from a DTM.
From the radiometric surveys we simply used the four variables produced by the surveyors (Geological Survey of Norway): ground concentration of Potassium, Thorium, Uranium, as well as total count.
From the DTMs we calculated several land surface parameters, ranging from simple terrain indices to more complex geomorphometric and hydrological variables \citep{maxwellLandsurfaceParametersSpatial2022}.
The categorical predictor was peat depth class, from a national map dataset.
Complete descriptions of predictors follow below.

@@

\caption{\label{tab:preds}Candidate predictors of peat depth.}
\centering
\begin{tabular}[t]{l|l}
\hline
name & description\\
\hline
radK & Potassium ground concentration\\
\hline
radTh & Thorium ground concentration\\
\hline
radU & Uranium ground concentration\\
\hline
radTC & Total count of gamma radiation\\
\hline
elevation & Mean elevation\\
\hline
slope1m & Mean of 1 m slope\\
\hline
TPI1m & Mean of 1 m topographic position index\\
\hline
TRI1m & Mean of 1 m terrain ruggedness index\\
\hline
roughness1m & Mean of 1 m roughness\\
\hline
slope10m & 10 m slope\\
\hline
TPI10m & 10 m topographic position index\\
\hline
TRI10m & 10 m terrain ruggedness index\\
\hline
roughness10m & 10 m roughness\\
\hline
MRVBF & Multi-resolution valley bottom flatness\\
\hline
TWI5m & Mean of 5 m topographic wetness index\\
\hline
TWI10m & 10 m topographic wetness index\\
\hline
TWI20m & Bilinear interpolation of 20 m topographic wetness index\\
\hline
TWI50m & Bilinear interpolation of 50 m topographic wetness index\\
\hline
DTW2500 & Depth-to-water index, flow initiation area of 0.25 ha\\
\hline
DTW5000 & Depth-to-water index, flow initiation area of 0.5 ha\\
\hline
DTW10000 & Depth-to-water index, flow initiation area of 1 ha\\
\hline
DTW20000 & Depth-to-water index, flow initiation area of 2 ha\\
\hline
DTW40000 & Depth-to-water index, flow initiation area of 4 ha\\
\hline
DTW80000 & Depth-to-water index, flow initiation area of 8 ha\\
\hline
DTW160000 & Depth-to-water index, flow initiation area of 16 ha\\
\hline
DMK & DMK peat depth class, categorical with 3 levels\\
\hline
\end{tabular}
\end{table}

\subsubsection{Radiometric}

@@

\subsubsection{Terrain}

@@

We used the \emph{terra} R package to calculate from the DTMs: slope, topographic position index (difference from mean of eight neighbors), terrain ruggedness index (mean of absolute differences from eight neighbors), and roughness (range in the nine-cell neighborhood).
These variables were derived at two scales to produce different predictors; we either calculated from \unit{1\,m} DTM resolution and then aggregated, or aggregated to \unit{10\,m} DTM resolution and then calculated the indices.
This kind of multiscale feature engineering of land surface parameters has been found to improve machine learning predictions of soil properties \citep{millerImpactMultiscalePredictor2015, dornikOptimalScalingPredictors2022, newmanAssessingSpatiallyHeterogeneous2023}.
We know that peat depth tends to vary at fine scales in Norway, which is why we chose \unit{1\,m} and \unit{10\,m} resolutions \citep{maxwellLandsurfaceParametersSpatial2022}.
We also calculated the the Multi-Resolution Valley Bottom Flatness index, which indicates the degree of valley bottom flatness at a given location via a multiscale algorithm \citep{gallantMultiresolutionIndexValley2003}.
We calculated this index in SAGA GIS \citep[v.9.3.2,][]{conradSystemAutomatedGeoscientific2015} with default parameters.

@@

\subsubsection{Peat depth class}

We prepared one categorical predictor --- peat depth class --- from a historical national map dataset called \emph{DMK} \citep{ahlstromAR5Klassifikasjonssystem2019}.
The DMK peat depth classes are: \textless{} \unit{1\,m} (\emph{shallow}), \textgreater{} \unit{1\,m} (\emph{deep}), and \emph{unknown}.
This dataset stems from field measurements made in 1964--2007 as part of a wider land cover mapping in Norway.
The mapping was primarily for identifying agricultural and silvicultural resources, so it covers productive areas below the tree line and has greater coverage of peat depth class in peatlands judged to be arable or afforestable \citep{bjordalMarkslagsklassifikasjonOkonomiskKartverk2007, ahlstromAR5Klassifikasjonssystem2019}.
Mappers generally assigned peat depth classes to polygons of at least \unit{0.5\,ha}, although delineating polygons down to \unit{0.2\,ha} was allowed if peat depth showed a ``particularly marked difference'' \citep{bjordalMarkslagsklassifikasjonOkonomiskKartverk2007}.
We rasterized the peat depth class attribute to our \unit{10\,m} grid.

@@

\subsubsection{Modelling approach}

@@

We implemented models in the \emph{tidymodels} framework in R \citep{kuhnTidymodelsCollectionPackages2020}, with the \emph{ranger} R package for RFs \citep[v.0.16,][]{wrightRangerFastImplementation2017}.
RFs were fit with 1000 trees, minimum node size of 5, and the number of predictors randomly sampled at each split was the square root of the total number of predictors (\emph{ranger} default).
We did not tune these hyperparameters because RFs are relatively insensitive to tuning \citep{probstHyperparametersTuningStrategies2019}, and because it would require nested spatial cross-validation to prevent data leakage \citep{schratzHyperparameterTuningPerformance2019}.

\subsubsection{Model performance}

For both sites we compared the performance of models with five different configurations of three groups of predictors.
Specifically, we trained models with:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  only DMK peat depth class (2 binary indicator variables from 3 levels)
\item
  only terrain (21 continuous variables)
\item
  terrain and DMK peat depth class (23 variables)
\item
  radiometric and terrain predictors (25 continuous variables)
\item
  all predictors (26 variables)
\end{enumerate}

These different configurations simulate different scenarios of data availability that are common in Norway or internationally.
Comparing the different configurations allowed us to isolate the added value of each of the predictor groups.
We did not explore configurations comprising radiometrics without terrain, because Lidar terrain surveys typically precede airborne radiometric surveys.
The models with only DMK peat depth class were simple linear models rather than RFs, and served to provide a fair comparison between the accuracy of the RF models and the existing national map of peat depth, calibrated on the same data.

The performance of digital soil mapping must be evaluated with reference to a specific purpose \citep[i.e., map vs.~model validation, interpolation vs.~extrapolation,][]{robertsCrossvalidationStrategiesData2017, milaNearestNeighbourDistance2022}, and here we aimed to evaluate maps of peat depth across the study areas.
In the absence of additional field work to collect a design-based independent validation set, we used a spatial cross-validation scheme to evaluate model performance \citep{wadouxSpatialCrossvalidationNot2021, meyerMachineLearningbasedGlobal2022}.
Specifically, we used k-Means Nearest Neighbor Distance Matching (kNNDM), which creates cross-validation folds that mimic the spatial prediction task that is defined as the goal \citep{linnenbrinkKNNDMCVKfold2024}.
In particular, kNNDM looks for the spatial assignment of training data to folds that minimizes the difference between two distributions: nearest neighbor distances between training and test locations in the cross-validation, and nearest neighbor distances between training and prediction locations for the model.
That way, the spatial separation between folds is similar to the separation between training and prediction locations --- which increases the quality of the map accuracy estimate \citep{linnenbrinkKNNDMCVKfold2024}.
For spatially clustered training data, this approach strikes a balance between the risk of optimistic metrics from random cross-validation and the risk of pessimistic metrics from other forms of spatial cross-validation \citep{wadouxSpatialCrossvalidationNot2021}.
We implemented the kNNDM with the \emph{CAST} R package \citep[v.1.0.2,][]{meyerCASTPackageTraining2024}, setting prediction locations to all AR5 mire cells in the study area, and choosing a number of folds (5--20) that produced the best match between the two NND distributions.
From the cross-validation we quantified \emph{Root Mean Squared Error} (accuracy, original scale), \emph{R\textsuperscript{2}} (correlation, standardized scale), and \emph{Lin's concordance correlation coefficient} (accuracy and correlation, standardized scale).

Uncertainty.

Extrapolation.

\subsubsection{Model interpretation}

We quantified global variable importance and examined partial dependence plots for the best-performing predictor configuration at each site.
Global variable importance measures the influence of a given predictor on the output of the model, aggregated across all locations.
Partial dependence plots depict the shape of the fitted relationship between a given predictor and the outcome.
Both are useful for understanding the mechanisms behind the model's predictions and the roles of the predictors in the model.

For both sites, we interpreted a model trained on a non-collinear subset of variables from the best performing predictor configuration --- because correlation between predictors degrades variable importance measures \citep{stroblConditionalVariableImportance2008, biauRandomForestGuided2016} and can produce misleading visualizations of predictor-outcome relationships \citep{biecekExplanatoryModelAnalysis2021, dwivediExplainableAIXAI2023}.
Specifically, we eliminated variables from the best performing predictor configuration to obtain a set with no pairwise Pearson correlation above 0.7.
Thus, highly-correlated sets of predictors are represented by a single variable for the purposes of model interpretation.

We calculated variable importance with the \emph{vip} R package (v.0.4.1), by three different methods: \emph{FIRM}, \emph{permutation}, and \emph{Shapley} \citep{greenwellVariableImportancePlots2020}.
\emph{FIRM} values measure the flatness of the partial dependence plot, \emph{permutation} values measure the decrease in model performance when the predictor is permuted, and \emph{Shapley} values are aggregated from local, game-theoretical measures of variable importance \citep{greenwellVariableImportancePlots2020}.
\emph{Permutation} values were obtained from ten iterations, with RMSE as the performance measure.

We calculated partial dependence with the \emph{pdp} R package \citep[v.0.8.1,][]{greenwellPdpPackageConstructing2017}.
For the top six most important variables, we plotted both partial dependence and individual conditional expectation (ICE), to show the average effect of the predictor on the outcome and the variation in the effect across observations, repectively \citep{goldsteinPeekingBlackBox2015}.
Non-parallel ICE lines indicate the presence of interactions between predictors.
